Parallel-in-time methods for Particle Simulations

### 2024-12-01 paper/main.tex
Paper

\begin{IEEEkeywords}
parareal, particle-simulations, parallel-in-time.
\end{IEEEkeywords}

I am greatful to my advisor, Samuel J. Newcome, M. Sc., for suggesting directions and providing valueable feedback.

@@

\subsection{The particle simulation}
We use a simple $n$-body particle simulation of the solar system with time complexity $\mathcal{O}(n^2)$, which we then integrate using the parareal method. Thus, we are integrating the following function:
$$
f(t, \mathbf{u}(t)) = f(t, (\mathbf{x}(t), \mathbf{v}(t))) = (\mathbf{v}(t), \mathbf{a}(t))
$$
where 
$$
a_i(t) = -G \sum_j m_j\frac{x_i(t)-x_j(t)}{\|x_i(t)-x_j(t)\|_2^3}
$$
where $m_j$ is the mass of the $j$-th particle. Acceleration is derived solely from the Newton's law of universal gravitation.

@@

\section{Implementation}

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal on 32 CPUs. For more information, see table \ref{runtimes}.}
\label{sim}
\end{figure}

@@

\begin{table*}[htbp]
\caption{Comparison of running times taken by different configurations of the simulator}
\begin{center}
\begin{tabular}{ccrrrrrrrrcc}
\toprule
Program & Time &  n CPUs & Segments & $N_\text{Coarse}$ & $N_{\text{Fine}}$ & $N_{\text{Final}}$ $^{\mathrm{a}}$ & Iters $^{\mathrm{b}}$ & Abs. error $^{\mathrm{c}}$ & Rel. error $^{\mathrm{d}}$ & CPU time & Runtime \\
\midrule
parareal & $10^4\text{days}$ & 32 & 32 &  5000 &  50000 & 1600000 & 1 & 0.0030 AU & 0.97\% & 00:05:33 & 00:00:35 \\
parareal & $10^4\text{days}$ & 16 & 16 & 10000 & 100000 & 1600000 & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\
serial & $10^4\text{days}$ & 1 & 1 & N/A & 1600000 & 1600000 & N/A & 0 AU & 0\% & 00:02:34 & 00:02:34 \\
\bottomrule
\multicolumn{11}{l}{$^{\mathrm{a}}$ Resulting number of steps (equal to number of segments times $N_\text{Fine}$)} \\
\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}$)} \\
\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation}
\end{tabular}
\label{runtimes}
\end{center}
\end{table*}

\begin{thebibliography}{00}
\bibitem{parareal}Bal, G. \& Maday, Y. A “Parareal” Time Discretization for Non-Linear PDE's with Application to the Pricing of an American Put. {\em Recent Developments In Domain Decomposition Methods}. pp. 189-202 (2002)
6-578 (2007), [Online]. Available: https://doi.org/10.1007/978-3-642-56118-4\_12
\bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)
\bibitem{parallelsolar}Saha, P., Stadel, J. \& Tremaine, S. A Parallel Integration Method for Solar System Dynamics. {\em The Astronomical Journal}. \textbf{114} pp. 409 (1997,7), http://dx.doi.org/10.1086/118485
\bibitem{parareal2}Gander, M. \& Vandewalle, S. ``Analysis of the Parareal Time‐Parallel Time‐Integration Method.'' SIAM Journal On Scientific Computing. 29, 556-578 (2007), [Online]. Available: https://doi.org/10.1137/05064607X
\end{thebibliography}

>>>>

\begin{IEEEkeywords}
parareal, particle simulations, parallel-in-time, solar system.
\end{IEEEkeywords}

@@

\subsection{The particle simulation}
We use a simple $n$-body particle simulation of the solar system with time complexity $\mathcal{O}(n^2)$, which we then integrate using the parareal method. This is inefficient for a large number of particles, however, our focus is on the parallel-in-time integration, therefore we use a simpler simulation.

Thus, we are integrating the following equation:
$$
\diff{\mathbf{u}(t)}{t} = f(t, \mathbf{u}(t)) = f(t, (\mathbf{x}(t), \mathbf{v}(t))) = (\mathbf{v}(t), \mathbf{a}(t))
$$
where 
$$
a_i(t) = -G \sum_j m_j\frac{x_i(t)-x_j(t)}{\|x_i(t)-x_j(t)\|_2^3}
$$
where $m_j$ is the mass of the $j$-th particle. Acceleration is derived solely from the Newton's law of universal gravitation.

It is also possible to use a similar algorithm to the parareal algorithm using a simplified model instead of the coarse solver. This was done prior to the invention of parareal by \cite{parallelsolar} also in a solar system simulation, who obtained the coarse approximation by simulating only the interactions of Sun and other planets.\cite{parallelsolar}\cite{farfuture}

@@

\subsection{Stopping criteria}
We assume the algorithm has converged when the maximum distance between the states of two consecutive iterations is smaller than some $\varepsilon$. We use $\varepsilon = 10^{-3}\text{ AU}$.

@@

\section{Implementation and experiments}
Two implementations have been created, first a reference implementation in Python and then a faster one in C++ with OpenMP.

In case we would want to use a substantially higher number of CPUs, we would have to use a different framework, such as MPI. However, due to the higher complexity and difficulty of such an implementation, an easier approach using OpenMP was used in this project for the present time instead.

Multiple experiments were ran on a cluster in order to get a more reproducible environment. The resulting run times were collected in Table \ref{runtimes}. For visualizations of the simulation results see Figures \ref{sim} and \ref{energy}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal on 32 CPUs. For more information, see Table \ref{runtimes}.}
\label{sim}
\end{figure}

@@

\begin{table*}[htbp]
\caption{Comparison of running times taken by different configurations of the simulator. 
$N_{\text{Coarse}}$ and $N_{\text{Fine}}$ refer to the number of steps of the coarse and the fine integrators respectively. 
To allow comparison with the classic sequential algorithm using only the fine integrator, $N_{\text{Final}}$ shows the resulting number of steps (for parareal it is equal to the number of segments times $N_\text{Fine}$).
\newline The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
\begin{center}
\begin{tabular}{ccrrrrrrrrcc}
\toprule
Program & Time &  n CPUs & Segments & $N_\text{Coarse}$ & $N_{\text{Fine}}$ & $N_{\text{Final}}$ & Iters $^{\mathrm{a}}$ & Abs. error $^{\mathrm{b}}$ & Rel. error $^{\mathrm{c}}$ & CPU time & Runtime \\
\midrule
parareal & $10^4\text{days}$ & 32 & 32 &  $5\times10^3$ &  $5\times10^4$ & $16\times10^5$ & 1 & 0.0030 AU & 0.97\% & 00:05:33 & 00:00:35 \\
parareal & $10^4\text{days}$ & 16 & 16 & $1\times10^4$ & $1\times10^5$ & $16\times10^5$ & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\
serial & $10^4\text{days}$ & 1 & 1 & N/A & $16\times10^5$ & $16\times10^5$ & N/A & 0 AU & 0\% & 00:02:34 & 00:02:34 \\
\bottomrule
\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation} \\
\end{tabular}
\label{runtimes}
\end{center}
\end{table*}

\section{Summary and discussion}

I would like to thank my advisor, Samuel J. Newcome, M. Sc., for suggesting directions and providing valueable feedback.

\begin{thebibliography}{00}
\bibitem{parareal}Bal, G. \& Maday, Y. A “Parareal” Time Discretization for Non-Linear PDE's with Application to the Pricing of an American Put. {\em Recent Developments In Domain Decomposition Methods}. pp. 189-202 (2002)
6-578 (2007), [Online]. Available: https://doi.org/10.1007/978-3-642-56118-4\_12
\bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)
\bibitem{parallelsolar}Saha, P., Stadel, J. \& Tremaine, S. A Parallel Integration Method for Solar System Dynamics. {\em The Astronomical Journal}. \textbf{114} pp. 409 (1997,7), http://dx.doi.org/10.1086/118485
\bibitem{farfuture}Gander, M. Is it possible to predict the far future before the near future is known accurately?. {\em Snapshots Of Modern Mathematics From Oberwolfach; 2019}. pp. 21 (2019), [Online]. Available: http://publications.mfo.de/handle/mfo/3693
\bibitem{parareal2}Gander, M. \& Vandewalle, S. ``Analysis of the Parareal Time‐Parallel Time‐Integration Method.'' SIAM Journal On Scientific Computing. 29, 556-578 (2007), [Online]. Available: https://doi.org/10.1137/05064607X
\end{thebibliography}
